Formalized Linear Programming

### 2020-01-03 TeX/paper.tex
add to paper

\begin{document}
\bibliographystyle{plain}
\lstset{language=lean}
\lstset{basicstyle=\normalsize}
\maketitle
\begin{abstract}
% What is the general topic we look at?
Correct and efficient linear programming, using the simplex
algorithm, is of high importance not only in the context of
operational research but also in programming language analysis.
% What is the specific research problem we face?
While existing solvers are robust, the lack of formally verified
implementations prevents their use in computational proofs and
rare corner-case bugs make the use of linear programming for
automatic program generation difficult.
% Why has nobody else adequately answered the research problem yet?
Today, existing implementations are either written in low-level
high-performance languages (e.g., C++), which are hard to verify, or
focus on the verification aspects but are not intended to be executable in practice.
% What is our key idea to address the stated research problem?
We address this problem by implementing an executable and verified
simplex solver in the Lean programming language.
% What are the key steps needed to successfully implement our idea?
We ensure both ease of verification and reasonable execution
performance by uses Bland's rule to select a pivot and by stating our
pivot function without large and costly matrix multiplications as
they are often used in textbook formalizations.
% What is the key global impact we envision our research to have?
Our research provides core foundations for verified linear
programming and demonstrates the practicality of Lean as a language
for fast verified constraint programming.

@@

The choice of pivot row and column satisfying (\ref{eqn:valid_pivot}) is non-unique, and must be taken carefully in order to guarantee termination. Different implementations of the simplex algorithm use different pivot rules to guarantee termination. If $z$ was strictly increasing it would be straightforward to prove the algorithm terminates, since there are only finitely many partitions of the variables into row and column variables, and the simplex algorithm cannot repeat a partition if $z$ is strictly increasing. However when $b_i = 0$ in the pivot row $i$, $z$ does not increase, and ensuring termination is more difficult. Our implementation uses Bland's Rule~\cite{Blands_rule} to ensure termination.

@@

>>>>

\begin{document}
\bibliographystyle{plain}
\lstset{language=lean}
\lstset{basicstyle=\normalsize}
\maketitle
\chris{We should mention somewhere that we only did Phase 2 of the textbook simplex}
\begin{abstract}
% What is the general topic we look at?
Correct and efficient linear programming, using the simplex
algorithm, is of high importance not only in the context of
operational research but also in programming language analysis.
% What is the specific research problem we face?
While existing solvers are robust, the lack of formally verified
implementations prevents their use in computational proofs and
rare corner-case bugs make the use of linear programming for
automatic program generation difficult.
% Why has nobody else adequately answered the research problem yet?
Today, existing implementations are either written in low-level
high-performance languages (e.g., C++), which are hard to verify, or
focus on the verification aspects but are not intended to be executable in practice.
% What is our key idea to address the stated research problem?
We address this problem by implementing an executable and verified
simplex solver in the Lean programming language.
% What are the key steps needed to successfully implement our idea?
We ensure both ease of verification and reasonable execution
performance by uses Bland's rule to select a pivot and by stating our
pivot function without large and costly matrix multiplications as
they are often used in textbook formalizations.
% What is the key global impact we envision our research to have?
Our research provides core foundations for verified linear
programming and demonstrates the practicality of Lean as a language
for fast verified constraint programming.

@@

The choice of pivot row and column satisfying (\ref{eqn:valid_pivot}) is non-unique, and must be taken carefully in order to guarantee termination. Different implementations of the simplex algorithm use different pivot rules to guarantee termination. If $z$ was strictly increasing it would be straightforward to prove the algorithm terminates, since there are only finitely many partitions of the variables into row and column variables, and the simplex algorithm cannot repeat a partition if $z$ is strictly increasing. However when $b_i = 0$ in the pivot row $i$, $z$ does not increase, and ensuring termination is more difficult. Our implementation uses Bland's Rule \cite{Blands_rule} to ensure termination.

@@

There are existing formalizations of the simplex algorithm in both Coq \cite{DBLP:journals/corr/AllamigeonK17} and Isabelle \cite{10.1007/978-3-642-32759-9_35}. Additionally the Verified Polyhedron Library \cite{VPL} has some formalization related to the simplex algorithm.

The Coq implementation differs from this implementation in that it was not intended to produce code that could be compiled into a reasonably fast executable algorithm but to use the simplex algorithm as a "tool" to be used to prove other theorems. Consequently they took a very different approach to the proof of the algorithm. They used symbolic perturbation as opposed to Bland's rule for the pivot rule, which leads to a more intuitive proof of termination. The approach taken to their proofs is very different
